Parameter Recovery Validation

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): distinguish model descriptions from those to which we apply predictive checks
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Parameter Recovery Validation
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: mathjax
    include-in-header: ../../mathjax.html
format-links: [ipynb]
execute:
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
author:
  - name: Pyrovelocity Team
    email: team@pyrovelocity.net
abstract: |
  This notebook demonstrates parameter recovery validation for PyroVelocity models using the direct dimensionless approach,
  progressing from analytical solutions with observed time to numerical simulation with latent time coordinates.
  We focus on models that naturally avoid identifiability issues while maintaining biological interpretability,
  using hierarchical priors on characteristic scales and lumped effective capture efficiency.
  We illustrate how to systematically validate that probabilistic models can recover
  known parameters from synthetic data, which is essential before applying models
  to real-world data where true parameter values are unknown.
keywords: [single-cell transcriptomics, probabilistic modeling, model calibration, model validation, RNA velocity]
bibliography: ../../references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

@@

This notebook demonstrates parameter recovery validation for PyroVelocity models using the dimensionless parameterization that naturally avoids identifiability issues while maintaining biological interpretability. We implement hierarchical priors on characteristic scales and lumped effective capture efficiency, progressing from analytical solutions with observed time to numerical simulation with latent time coordinates. For each model we include a graphical representation of the model's generative process as well as a mathematical description of the prior, observation model, and likelihood.

## Model Progression

We validate parameter recovery across a progression of **dimensionless models** that naturally avoid identifiability issues while maintaining biological interpretability. Each model builds complexity while preserving the core dimensionless parameterization.

@@

# Set random seeds for reproducibility
torch.manual_seed(42)
pyro.set_rng_seed(42)
np.random.seed(42)

print("Setup complete!")
```

@@

**Mathematical Prohibition Against Parameter Averaging**: The posterior predictive distribution is **NOT** computed using averaged parameters:

$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)} \quad \text{(MATHEMATICALLY INCORRECT)}$$

Using $\bar{\theta}$ violates the fundamental definition of posterior predictive distributions and fails to capture parameter uncertainty propagation through the nonlinear piecewise dynamics.

**Correct Sample-wise Generation**: Each $\tilde{y}^{(s)}$ must be generated using the complete parameter vector $\theta^{(s)}$:

@@

@@

The mathematical definition of posterior predictive distributions explicitly requires sample-wise generation. **Parameter averaging is mathematically incorrect** because:

1. **Nonlinear dynamics**: The piecewise analytical solutions are nonlinear functions of parameters, so $f(\mathbb{E}[\theta]) \neq \mathbb{E}[f(\theta)]$
2. **Phase discontinuities**: Averaged temporal parameters may place cells in incorrect phases
3. **Uncertainty underestimation**: Averaging eliminates parameter uncertainty propagation
4. **Invalid inference**: Results do not represent valid Bayesian posterior predictive distributions

>>>>

---
title: Parameter Recovery Validation
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: mathjax
    include-in-header: ../../mathjax.html
format-links: [ipynb]
execute:
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
author:
  - name: Pyrovelocity Team
    email: team@pyrovelocity.net
abstract: |
  This notebook demonstrates parameter recovery validation for PyroVelocity models using dimensionless models.
  We progress from analytically tractable models where the temporal coordinate is an observed variable toward the more complicated situation of models with latent variables that represent temporal ordering and must therefore be inferred.
  We focus on constructing models that naturally avoid identifiability issues while maintaining biological interpretability,
  using hierarchical priors on characteristic scales and simple accounting of technical artifacts through effective parameters that aggregate sources of error.
  We illustrate how to systematically validate that probabilistic models can recover
  known parameters from synthetic data, which is essential before applying models
  to real-world data where true parameter values are unknown.
keywords: [single-cell transcriptomics, probabilistic modeling, model calibration, model validation, RNA velocity]
bibliography: ../../references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

@@

This notebook demonstrates parameter recovery validation for PyroVelocity models using the dimensionless parameterization that naturally avoids identifiability issues while maintaining biological interpretability. We implement hierarchical priors on characteristic scales and lumped effective capture efficiency, progressing from analytical solutions with observed time to numerical simulation with latent time coordinates. For each model we include a graphical representation of the model's generative process as well as a mathematical description of the prior, observation model, and likelihood. We illustrate how to perform a somewhat simplified version of prior and posterior predictive checking for the piecewise activation model with latent time.

## Model Development

We work through a progression of **dimensionless models** that naturally avoid or aim to avoid identifiability issues while maintaining biological interpretability. Each model builds complexity while preserving the core dimensionless parameterization.

@@

# Set random seeds for reproducibility
torch.manual_seed(42)
pyro.set_rng_seed(42)
np.random.seed(42)
```

@@

**Mathematical Prohibition Against Parameter Averaging**:

**Correct Sampling**: Each $\tilde{y}^{(s)}$ must be generated using the complete parameter vector $\theta^{(s)}$:

@@

The posterior predictive distribution cannot be computed using averaged parameters such as

$$\bar{\theta} = \frac{1}{S}\sum_{s=1}^S \theta^{(s)}$$

Using $\bar{\theta}$ would violate the fundamental definition of posterior predictive distributions and fails to capture parameter uncertainty propagation through the nonlinear dynamics.

@@

The mathematical definition of posterior predictive distributions explicitly requires sampling parameters and observables jointly, without ad hoc averaging.
